Superphy: Real-time prediction of \textit{E. coli} phenotype using graph-based genome sequence storage and retrieval with the \textit{spfy} module

### 2017-11-27 paper-webserver.tex
FIX: mostly spelling

\subsection{Data Storage}
% para
% 0. Goals: big-data, everything linked, easy addition of new links
% 1. spfy is built around graph technologies
% 1. how we structure our graph
% 2. ontoogies used
% 3. inferencing
% 3 1/2. SPARQL queries
% para: the semantic web
Spfy is built around semantic web technologies which describe the relaions between different data \cite{berners2001semantic}. For biological data, individual data points such as genome, contiguous DNA sequence, or gene, are linked together in a queryable graph structure. This allows novel data to be seamlessly incorporated into the existing graph, and has the use of graph databases for relational information has been proposed as a common standard for the open sharing of data \cite{horrocks2005semantic}.

@@

% note: below is a feature in development
The permanent storage of results is as a one-time cost, and allows population-wide analyses of all stored genomes; result storage also enables Spfy to avoid recomputation when the same analysis is re-run. To identify analysis tasks, Spfy fingerprints requests by computing its SHA1 hash and tags the results in the graph database with this hash. Subsequent queries are then compared through fingerprinting, and results can be rebuilt from the graph by searching for matching fingerpirnts.

@@

% drawbacks - big data
Many previous bioinformatics software programs have been developed \textit{ad hoc}, with inividual researchers and laboratories developing software specific to their environment \cite{de2015trends}.
Such tools were often script-based, with custom data formats, and only suitable for small collections of data \cite{de2015trends}.
While this was acceptable for smaller analyses, bioinformatic pipelines utilizing WGS data are larger and involve linked dependencies, which require the application of systems engineering principles \cite{schatz2015biological}.
Additionally, many subsets of biology now require the analyses of big-data, where the ability to perform computations in real-time, store data in flexible databases, and utilize a common application programming interface (API) linking resources are required \cite{swaminathan2016review}.

One of the key goals in developing Spfy is to maintain instantaneity, as modern websites have accustomed users to immediate results.
We attempt to use innovations in web development to bring a similar experience to Spfy as a predictive genomics platform for \textit{E. coli}.
The use of Docker containerization, task queues, and a graph database, were all chosen to support \textit{instantaneity}; modern analytics platforms must respond to user-requested analyses of big-data in the same efficiancy as old analyses of single files.

@@

% para
% focus on application
The isolation and characterization of bacterial pathogens are critical for Public Health laboratories to rapidly respond to outbreaks, and to effectively monitor known and emerging pathogens through surveillance programs.
Until recently, public-health agencies relied on laboratory tests such as serotyping, pulsed-field gel electrophoresis (PFGE) banding patterns, or Polymerase Chain Reaction (PCR)-based amplication to identify known VFs or AMR genes, along with other tests \cite{ronholm2016navigating}, to characterize bacterial isolates in outbreak and surveillance settings.

@@

% application: results similar to a wet-lab
Spfy uses WGS results.
After intial sequencing of new isolates, Spfy can be used in place of a traditional reference laborartory, to determine the O-type and H-type, Stx type, and all known VFs and AMR genes in real-time.
These results can be shared with other ageencies and researchers over the internet.
Futhermore, using Spfy's database of pre-processed genomes, Spfy can determine all strains a sample may be related to which is useful for monitoring the evolution of pathogens over time.
A computational approach saves the time and cost associated with performing multiple tests per sample, and allows population comparisons which would not be possible with the current pace of WGS results.

@@

% namely galaxy
Other scientific workflow technologies such as Galaxy \cite{goecks2010galaxy}, and bioinformatics pipelines IRIDA and BAP are used to run analysis modules on WGS data.
Galaxy aims to provide a reproducible compuation-based research environment which is accessible to individuals without programming knowledge. Galaxy tackles the problem of linking different analysis software together, by defining interdependencies using a custom schema. A visual workflow editor is also provided for ease of use.
IRIDA is build on top of the Galaxy framework, and adds prebuilt pipelines specific to bioinformatics uses, as well as sequence and result storage. IRIDA takes a project-based approach, with sequences stored per project, and results store linearly per sequence. Focus is given on refined controls for collaborating on projects, and common terms found in the GenEpiO ontology are used to describe results.

@@

In Spfy, we take a more opionionated approach to result storage and data integration. The benefits of real-time analysis pipelines as found in BAP, and the use of the GenEpiO ontology as in IRIDA, are included in Spfy.
On a per file basis, Spfy performs at a similar speed to BAP on predictive genomics tasks, though Spfy does not provide genome assembly services.
Spfy also processes XXXX files over XX tasks in XXXX time by using a novel approach of distributing computations over a task queue and multiple Docker compartmentalized containers.
In place of a relational database for storing the location of result files, Spfy parses result data into graph objects and integrates results into a persistant graph database.
This enables Spfy to create large datasets where analysis results are linked to genomes, and to perform population-wide analyses. In all, Spfy provides similar functionallity with an expanded focus on integrated result storage and big-data analyses.

To approach the challenge of integrating different bioinformatics programs, Spfy instead uses technologies prevalent in common web services not necessarily related to scientific workflows.
While we prefer to package individual bioinformatics programs using language-agnostic package managers such as Conda \url{https://conda.io/}, we do agree that suites of bioinformatics software benefit from having all dependencies installed and deployable in Docker containers \cite{di2015impact}.
These programs are connected to form a workflow using task queues; a design methodology used to offload long-running or non-instantaneous tasks over available computing infrastructure.
Unlike platforms such as Galaxy which favor a drag-and-drop design for building workflows, Spfy favors a tightly coupled approach explicitedy linking together different bioinformatics programs within the task queue implemntation.
In particular, Spfy uses asynchronous tasks queues which are tripped in response to user requests, thereby allowing immediate responsivity to users on the website, and processes generating human-readable results can be separated from processes handling long-term result storage.

% standard web tech means you can deploy to different cloud computing services
One of the key benefits of using more common-place technologies is the compatibility with other infrastructure resources.
Docker containerization is widely supported by cloud computing services: Amazon Web Services (AWS) \url{https://aws.amazon.com/docker/}, Google Cloud Platform (GCloud) \url{https://cloud.google.com/container-engine/}, and Microsoft Azure \url{https://azure.microsoft.com/en-us/services/container-service/}, and self-hosted cloud computing technologies such as OpenStack \url{https://wiki.openstack.org/wiki/Docker}.
% we may want to sign up for a free trial to give an example of this
Spfy packages compute nodes, which manage a collection of task queue workers, as reproducible Docker containers which can be networked together to form a compute cluster.
Docker containerization has a neglible impact on performance \cite{di2015impact}, and allows the platform to easily scale to demand.
In contrast to Galaxy, Spfy is a targeted approach building a predictive analytics platform for \textit{E. coli} which leverages generic web technologies to favor a responsive, big-data design strategy tackling the challenge \cite{fricke2014bacterial} of integrating results from multiple genomes and examining shared connections.
By packaging the individual bioinformatics software as Conda packages, they can be easily ported to Galaxy.

>>>>

\subsection{Data Storage}
% para
% 0. Goals: big-data, everything linked, easy addition of new links
% 1. spfy is built around graph technologies
% 1. how we structure our graph
% 2. ontoogies used
% 3. inferencing
% 3 1/2. SPARQL queries
% para: the semantic web
Spfy is built around semantic web technologies which describe the relations between different data \cite{berners2001semantic}. For biological data, individual data points such as genome, contiguous DNA sequence, or gene, are linked together in a query-able graph structure. This allows novel data to be seamlessly incorporated into the existing graph, and has the use of graph databases for relational information has been proposed as a common standard for the open sharing of data \cite{horrocks2005semantic}.

@@

% note: below is a feature in development
The permanent storage of results is as a one-time cost, and allows population-wide analyses of all stored genomes; result storage also enables Spfy to avoid recomputation when the same analysis is re-run. To identify analysis tasks, Spfy fingerprints requests by computing its SHA1 hash and tags the results in the graph database with this hash. Subsequent queries are then compared through fingerprinting, and results can be rebuilt from the graph by searching for matching fingerprints.

@@

% drawbacks - big data
Many previous bioinformatics software programs have been developed \textit{ad hoc}, with individual researchers and laboratories developing software specific to their environment \cite{de2015trends}.
Such tools were often script-based, with custom data formats, and only suitable for small collections of data \cite{de2015trends}.
While this was acceptable for smaller analyses, bioinformatic pipelines utilizing WGS data are larger and involve linked dependencies, which require the application of systems engineering principles \cite{schatz2015biological}.
Additionally, many subsets of biology now require the analyses of big-data, where the ability to perform computations in real-time, store data in flexible databases, and utilize a common application programming interface (API) linking resources are required \cite{swaminathan2016review}.

One of the key goals in developing Spfy is to maintain instantaneity, as modern websites have accustomed users to immediate results.
We attempt to use innovations in web development to bring a similar experience to Spfy as a predictive genomics platform for \textit{E. coli}.
The use of Docker containerization, task queues, and a graph database, were all chosen to support \textit{instantaneity}; modern analytics platforms must respond to user-requested analyses of big-data in the same efficiency as old analyses of single files.

@@

% para
% focus on application
The isolation and characterization of bacterial pathogens are critical for Public Health laboratories to rapidly respond to outbreaks, and to effectively monitor known and emerging pathogens through surveillance programs.
Until recently, public-health agencies relied on laboratory tests such as serotyping, pulsed-field gel electrophoresis (PFGE) banding patterns, or Polymerase Chain Reaction (PCR)-based amplification to identify known VFs or AMR genes, along with other tests \cite{ronholm2016navigating}, to characterize bacterial isolates in outbreak and surveillance settings.

@@

% application: results similar to a wet-lab
Spfy uses WGS results.
After initial sequencing of new isolates, Spfy can be used in place of a traditional reference laboratory, to determine the O-type and H-type, Stx type, and all known VFs and AMR genes in real-time.
These results can be shared with other agencies and researchers over the internet.
Furthermore, using Spfy's database of pre-processed genomes, Spfy can determine all strains a sample may be related to which is useful for monitoring the evolution of pathogens over time.
A computational approach saves the time and cost associated with performing multiple tests per sample, and allows population comparisons which would not be possible with the current pace of WGS results.

@@

% namely galaxy
Other scientific workflow technologies such as Galaxy \cite{goecks2010galaxy}, and bioinformatics pipelines IRIDA and BAP are used to run analysis modules on WGS data.
Galaxy aims to provide a reproducible computation-based research environment which is accessible to individuals without programming knowledge. Galaxy tackles the problem of linking different analysis software together, by defining interdependencies using a custom schema. A visual workflow editor is also provided for ease of use.
IRIDA is build on top of the Galaxy framework, and adds prebuilt pipelines specific to bioinformatics uses, as well as sequence and result storage. IRIDA takes a project-based approach, with sequences stored per project, and results store linearly per sequence. Focus is given on refined controls for collaborating on projects, and common terms found in the GenEpiO ontology are used to describe results.

@@

In Spfy, we take a more opinionated approach to result storage and data integration. The benefits of real-time analysis pipelines as found in BAP, and the use of the GenEpiO ontology as in IRIDA, are included in Spfy.
On a per file basis, Spfy performs at a similar speed to BAP on predictive genomics tasks, though Spfy does not provide genome assembly services.
Spfy also processes XXXX files over XX tasks in XXXX time by using a novel approach of distributing computations over a task queue and multiple Docker compartmentalized containers.
In place of a relational database for storing the location of result files, Spfy parses result data into graph objects and integrates results into a persistent graph database.
This enables Spfy to create large datasets where analysis results are linked to genomes, and to perform population-wide analyses. In all, Spfy provides similar functionality with an expanded focus on integrated result storage and big-data analyses.

To approach the challenge of integrating different bioinformatics programs, Spfy instead uses technologies prevalent in common web services not necessarily related to scientific workflows.
While we prefer to package individual bioinformatics programs using language-agnostic package managers such as Conda \url{https://conda.io/}, we do agree that suites of bioinformatics software benefit from having all dependencies installed and deployable in Docker containers \cite{di2015impact}.
These programs are connected to form a workflow using task queues; a design methodology used to offload long-running or non-instantaneous tasks over available computing infrastructure.
Unlike platforms such as Galaxy which favor a drag-and-drop design for building workflows, Spfy favors a tightly coupled approach explicitly linking together different bioinformatics programs within the task queue implementation.
In particular, Spfy uses asynchronous tasks queues which are tripped in response to user requests, thereby allowing immediate responsivity to users on the website, and processes generating human-readable results can be separated from processes handling long-term result storage.

% standard web tech means you can deploy to different cloud computing services
One of the key benefits of using more common-place technologies is the compatibility with other infrastructure resources.
Docker containerization is widely supported by cloud computing services: Amazon Web Services (AWS) \url{https://aws.amazon.com/docker/}, Google Cloud Platform (GCloud) \url{https://cloud.google.com/container-engine/}, and Microsoft Azure \url{https://azure.microsoft.com/en-us/services/container-service/}, and self-hosted cloud computing technologies such as OpenStack \url{https://wiki.openstack.org/wiki/Docker} all support Docker.
% we may want to sign up for a free trial to give an example of this
Spfy packages compute nodes, which manage a collection of task queue workers, as reproducible Docker containers which can be networked together to form a compute cluster.
Docker containerization has a negligible impact on performance \cite{di2015impact}, and allows the platform to easily scale to demand.
In contrast to Galaxy, Spfy is a targeted approach building a predictive analytics platform for \textit{E. coli} which leverages generic web technologies to favor a responsive, big-data design strategy tackling the challenge \cite{fricke2014bacterial} of integrating results from multiple genomes and examining shared connections.
By packaging the individual bioinformatics software as Conda packages, they can be easily ported to Galaxy.
